---
title: "dinocyst database"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dinocystDB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  engine.opts = list(bash = "-l")
)
```

```{r setupRMD}
library(here)
i_am("vignettes/dinocyst-database.Rmd")
```

Setting-up a SQLite database for the IODP sample locations.

```{bash setupDB, eval = F}
sqlite3 
.open 'inst/extdata/transferice.sqlite' # open db
.nullvalue NULL
.headers on
.mode column
```

# Connect to database from R

```{r connect}
# path to database
dbpath <- fs::path_package(package = "transferice", "extdata", "transferice.sqlite")
con <- DBI::dbConnect(RSQLite::SQLite(), dbpath)
```

# Write tables to database

## Dinocyst count data

First we collect dinocyst count data from a number of external sources.

```{r count}
library(readxl) # reading excel
library(tibble) # tibble data structure
library(janitor) # cleaning variable names
library(dplyr) # data wrangling
library(tidyr) # tidying data

# metadata
meta_vc <- c("sample_age_ma", "sample_id", "longitude", "latitude", "source_id", 
             "site", "hole", "sample_depth_mbsf", "site_hole", 
             "source_citation")

# the sheets of the excel workbook
shts <- excel_sheets(here("data-raw/SurfaceDinos.xlsx"))

# read dinocyst data (Thole 2021)
dino_thole <- read_xlsx(
  here("data-raw/SurfaceDinos.xlsx"), 
  sheet = shts[1],
  # explicit type casting is important
  col_types = c("text", rep("numeric", 38))
) |> 
  # clean variable names
  clean_names() |>  
  # sufrace sample
  mutate(sample_depth_mbsf = 0) |> 
  add_column(source_citation = "Thole et al 2022") |> 
  select(-c(non_quat_dinos, depth, weight, lycopodium, lyc_tab), site_hole = station)

# read dinocyst data (Marret 2019)
dino_marret <- read_xlsx(
  here("data-raw/SurfaceDinos.xlsx"), 
  sheet = shts[3],
  # explicit type casting is important
  col_types = c("text", "numeric", "numeric", "text", rep("numeric", 92))
)  |>  
  # sufrace sample
  mutate(sample_depth_mbsf = 0) |> 
  # clean variable names
  clean_names() |> 
  select(-dino_tot) |> 
  rename(source_citation = contributors, site_hole = station)
```

Fossil data is extracted from the Pangaea database with `pangaear`.

```{r pangaear}
library(pangaear) # connect to the Pangaea database
library(purrr) # vectorised operations
library(stringr) # regex

# get data
dino_brinkhuis_raw <- pg_data("10.1594/PANGAEA.779869") 

# counts
dino_brinkhuis <- dino_brinkhuis_raw [[1]]$data |> 
  rename(
    source_id = `Sample label (DSDP/ODP/IODP sample designation)`, 
    sample_depth_mbsf = `Depth [m]`
  ) |> 
  # not needed info
  select(
    -c(
      `Acr (skolochorate; samples charact...)`, 
      `Dinofl cyst [#] (Counting, dinoflagellate cysts)`, 
      `Palyn ter [%]`, 
      `Palyn mar [%]`
    )
  )|> 
  mutate(
    sample_id = row_number(),
    longitude = 144.412700,
    latitude = -42.609680,
    sample_age_ma = `Age model [ka]` / 1000,
    source_citation = "Brinkhuis et al., 2003",
    .keep = "unused"
  ) |> 
  separate(source_id, into = c("site", "hole", "source_id"), sep = "-", 
           remove = FALSE, extra = "merge", fill = "right") |> 
  mutate(site_hole = paste(site, hole, sep = "-")) |>
  # clean dinocysts taxons
  rename(`Homotryblium oceancium` = `Hystrichokolpoma sp. [#] (cf. Homotryblium oceancium; o...)`) |>  # remove duplicated name Hystrichokolpoma sp. 
  rename_with(~str_remove(.x, "(?<=\\[#\\])(.)*")) |> 
  rename_with(~str_trim(str_remove(.x, "\\[#\\]")))
```

Recode taxon names

```{r species}

# read taxon list (Thole et al)
taxons_thole <- read_xlsx(
  here("data-raw/SurfaceDinos.xlsx"), 
  sheet = shts[4],
  col_names = c("x", "name", "abbrev", "other")
) 

# subspecies of Selenopemphix antarctica (manual) and clean names
taxons_thole <- mutate(
  taxons_thole, 
  abbrev = tolower(gsub("[^[:alnum:]]+", "", .data$abbrev)),
  name = if_else(.data$abbrev == "sant", "Selenopemphix antarctica sp1", .data$name),
  name = gsub("Cyst of ", "", .data$name)
) |> 
  add_row(abbrev = "sant_2", name = "Selenopemphix antarctica sp2") 

# manual addition from Marret
taxons_marret <- c(
  `Achomosphaera andalousiense` = "aand",
  `Caspidinium rugosum` = "crug",
  `Impagidinium caspienense` = "icas",
  `Impagidinium japonicum` = "ijap",
  `Lingulodinium machaerophorum` = "lmac",
  `Melitasphaeridium choanophorum` = "Mcho",
  `Nematosphaeropsis rigida` = "nrig",
  `Operculodinium aguinawense` = "oagu",
  `Operculodinium giganteum` = "ogig",
  `Spiniferites cruciformis` = "scru",
  `Spiniferites elongatus` = "selo", 
  `Spiniferites bentorii` = "sben",
  `Spiniferites sp. granular type` = "sgra",
  `Scrippsiella trifida` = "stri",
  `Echinidinium karaense` =  "ekar",
  `Trinovantedinium variabile` = "tvar",
  `Polykrikos sp. arctic morphotype` = "parc",
  `Polykrikos hartmannii` = "phar",
  `Echinidinium zonneveldiae` = "ezon",
  `Stelladinium bifurcatum` = "sbif",
  `Tuberculodinium vancampoae` = "tvan",
  `Biecheleria baltica` = "bbal",
  `Peridinium ponticum` = "ppon",
  `Protoperidinium monospinum` = "pmon",
  `Echinidinium bispiniformum` = "ebis",
  `Oblea acanthocysta` = "oaca",
  `Trinovantedinium pallidifulvum` = "tpal",
  `Protoperidium minutum` = "pmin",
  `Melitasphaeridium choanophorum` = "mcho"
)

# get full dinocyst names
taxons_brinkhuis <- dino_brinkhuis_raw[[1]]$metadata$parameters[11:134] |> 
  map_chr(list(1))

# filter for names and clean them
nms <- taxons_brinkhuis[str_detect(taxons_brinkhuis, "#")] |> 
  map(~str_split(.x, "\\[#\\]")) |> 
  flatten()

# abbreviations
abbr <- map_chr(nms, list(2)) |> 
  str_remove_all("[\\(\\)]") |> 
  str_trim()

clean_nms <- map_chr(nms, list(1)) |> 
  str_trim()

# all names combined in named vector
taxon_nms <- set_names(taxons_thole$abbrev, nm = taxons_thole$name) |> 
  append(set_names(taxons_marret, nm = names(taxons_marret))) |> 
  append(set_names(abbr, nm = clean_nms))
```

Now we are going to combine the data from these three sources

````{r all}
# combine both dinocyst count datasets
dino_all <- list(dino_marret, dino_thole, dino_brinkhuis) |> 
  map(~rename(.x, any_of(taxon_nms))) |> 
  bind_rows() |> 
  mutate(sample_id = row_number(), .before = "site_hole")

# remove all NA rows
dino_all <- filter(dino_all, !if_all(-any_of(meta_vc), ~is.na(.)))

# extract dino count data
dino_cnt <- select(dino_all, -any_of(meta_vc[meta_vc != "sample_id"]))

# extract dino names
dino_nms <- tibble(species = names(dino_cnt)[names(dino_cnt) != "sample_id"]) |> 
  mutate(taxon_id = row_number(), .before = "species")

# nsb long format
dino_cnt <- pivot_longer(
  dino_cnt, 
  -sample_id, 
  names_to = "taxon_id", 
  values_to = "taxon_abundance", 
  values_drop_na = TRUE
) |> 
  # code taxon id
  mutate(
    taxon_id =  
      recode(
        taxon_id, 
        !!!set_names(dino_nms$taxon_id, nm = dino_nms$species)
      )
  )

# to database (NSB neptune_sample_taxa)
DBI::dbWriteTable(con, "neptune_sample_taxa", dino_cnt, overwrite = TRUE)
```

# taxon list

```{r taxonlist}
# recode taxons
dino_nms <- tidyr::separate(
   dino_nms,
   species, 
   sep = " ", 
   into = c("genus", "species", "add"),  
   extra = "merge", 
   fill = "right"
  ) |> 
   tidyr::separate(
    add, 
    sep = "sensu", 
    into = c("subspecies", "species_author"),
    extra = "merge", 
    fill = "right"
  ) 

# write table to DB (NSB neptune_taxonomy)
DBI::dbWriteTable(con, "neptune_taxonomy", dino_nms, overwrite = TRUE)
```

# Locations

```{r locations}
# extract geographic information
dino_loc <- select(dino_all, site_hole, site, hole, longitude, latitude) |> 
  distinct(site_hole, .keep_all = TRUE) |> 
  mutate(hole_id = row_number(), .before = "site_hole")

# to database (NSB neptune_hole_summary)
DBI::dbWriteTable(con, "neptune_hole_summary", dino_loc, overwrite = TRUE)
```

# Study metadata

```{r sample}
# sample metadata selected
dino_smp <- select(
  dino_all, 
  any_of(c("sample_id", meta_vc[!meta_vc %in% c("longitude", "latitude",  "site",  "hole")]))
) |> 
  # code hole_id
  mutate(
    hole_id =  
      recode(
        site_hole, 
        !!! set_names(dino_loc$hole_id, nm = dino_loc$site_hole),
      )
  ) |> 
  select(-site_hole) # drop station/hole

# metadata
dino_sty <- distinct(dino_smp, source_citation, source_id) |> 
  mutate(dataset_id = row_number(), .before = "source_citation") 

# write table to DB
DBI::dbWriteTable(con, "neptune_sod", dino_sty)
```


# Sample metadata

The ..

```{r units, eval = FALSE}
# code dataset_id
dino_smp <- mutate(
  dino_smp,
  dataset_id =  
    recode(
      source_citation, 
      !!! set_names(dino_sty$dataset_id, nm = dino_sty$source_citation),
    )
) |> # lose contributors
  select(-c(source_citation, source_id))

# to database (NSB neptune_sample)
DBI::dbWriteTable(con, "neptune_sample", dino_smp, overwrite = TRUE)
```


Useful sources working with SQLite and DBconnect

- https://www.datacamp.com/community/tutorials/beginners-guide-to-sqlite
- https://db.rstudio.com/databases/sqlite/
- https://db.rstudio.com/getting-started/database-queries/